Empirical Analysis of Performance Bottlenecks in Graph Neural Network Training and Inference with GPUs

### 2020-12-26 NEUROC/Response-MajRev/response_letter-version2.tex

\section*{Response to the Reviewers}
% General intro text goes here
We thank the reviewers for their critical assessment and insightful comments on our work.
%
We have made extensive modifications to our manuscript, including adding more experiments and improving the presentation.
%
In the following, we first summarize the major changes we made.
%
%We also prepare an \emph{annotated version} for our revised manuscript.
%
%In the annotated version, changes corresponding to each point are all highlighted by red squares.

\subsection*{Summary of Changes}

\begin{enumerate}
\item According to reviewer 1's comments, we have added experiments on the performance bottleneck analysis of GNN inference.
%
Specifically, we added experiments on how the hyper-parameters of GNNs affect the elapsed time and the memory usage of inference in Section 4.1.2 ``Effects on Inference Time'' and Section 4.1.3 ``Effects on Peak Memory Usage'', respectively.
%
We presented the time breakdown analysis of GNN inference in Section 4.2.4.
%
We also presented the memory usage analysis of GNN inference in Section 4.3., and the performance bottlenecks of sample-based inference were analyzed in Section 4.4.2.

\item According to reviewer 2's comments, we have added an evaluation of how the model complexity and the sampling techniques affect the accuracy of GNNs.
%
Specifically, we evaluated the effects of hyper-parameters of GNNs on accuracy in Section 4.1.4.
%
We also compared the accuracy of the four typical GNNs in that section.
%
Finally, we analyzed the effects of the sampling techniques on accuracy in Section 4.4.4.

### 2020-12-26 NEUROC/Response-MajRev/response_letter-version2.tex

\subsection*{Summary of Changes}

\begin{enumerate}
\item According to reviewer 1's comments, we have added experiments on the performance bottleneck analysis of GNN inference.
%
Specifically, we added experiments on how the hyper-parameters of GNNs affect the elapsed time and the memory usage of inference in Section 4.1.2 ``Effects on Inference Time'' and Section 4.1.3 ``Effects on Peak Memory Usage'', respectively.
%
We presented the time breakdown analysis of GNN inference in Section 4.2.4.
%
We also presented the memory usage analysis of GNN inference in Section 4.3., and the performance bottlenecks of sample-based inference were analyzed in Section 4.4.2.

\item According to reviewer 2's comments, we have added an evaluation of how the model complexity and the sampling techniques affect the accuracy of GNNs.
%
Specifically, we evaluated the effects of hyper-parameters of GNNs on accuracy in Section 4.1.4.
%
We also compared the accuracy of the four typical GNNs in that section.
%
Finally, we analyzed the effects of the sampling techniques on accuracy in Section 4.4.4.

\item We have adjusted the abstract, Section 1 ``Introduction'', Section 5 ``Insights'', and Section 7 ``Conclusion'' according to the new findings on inference time and accuracy.

@@

\begin{point}
    Hope authors supplement the effect of sampling and GNNs on accuracy while comparing different complexity of model and sampling methods.
\end{point}

\begin{reply}

    We are grateful for your insightful suggestion.
    %
    Followed your suggestion, we additionally evaluate the effects of model complexity on accuracy in Section 4.1 ``Effects of Hyper-parameters on Performance'' in the revised manuscript.
    %
    We add discussion on how the hyper-parameters of GNNs affect the accuracy.
    %
    We also compare the test accuracy of different GNN models on the same dataset.

    We also additionally evaluate the effects of sampling techniques on accuracy in Section 4.4 ``Effects of Sampling Techniques'' in the revised manuscript.
    %
    We present the experimental results and related discussion in a new subsection Section 4.4.4 ``Effects on Accuracy''.
\end{reply}

### 2020-12-26 NEUROC/Response-MajRev/response_letter-version2.tex

    \end{figure}

    Second, the relative accuracy between the four GNNs varies greatly with different datasets.
    %
    There is no absolute winner.
    %
    We compare the best accuracy that each GNN achives on every dataset in Figure~\ref{fig:exp_hyperparameter_on_accuracy_alg_contrast}.
    %
    The best accuracy of the four GNNs are very close.
    %
    It is also close to the accuracy reported in [\cite{shchur2018_pitfall_of_gnn, zeng2020_graphsaint}].
    %
    GaAN that has the highest complexity performs slightly better than the other GNNs, winning three out of five datasets.
    %
    Simple GNN models (such as GCN) can still achieve good accuracy with proper hyper-parameter settings.

    \begin{figure}[h]
        \centering
        \includegraphics[width=3.5in]{../figs/experiments/exp_hyperparameter_on_accuracy_alg_contrast.pdf}
        \caption{Best accuracy that each GNN achieves on different datasets.}
        \label{fig:exp_hyperparameter_on_accuracy_alg_contrast}
    \end{figure}

    Besides, we add more details on the proportions of the train/evaluation/test sets of the datasets in Section 3.1 ``Experimental Setting'' in the revised manuscript.
    %
    Since the vertices of the \texttt{cam} dataset are not attached with feature vectors, we exclude it from the accuracy evaluation.
\end{reply}

@@

    The results indicate that the relationships between batch size and test accuracy are complex, especially for the neighbor sampler.
    %
    Larger batch size does not always bring higher accuracy, like GAT in Figure~\ref{fig:exp_graphsage_sampling_accuracy_on_fli}. 
    %
    Our observations are similar to [\cite{zeng2020_graphsaint}].
    %
    We believe that how to automatically select a proper batch size is a topic worth studying.
\end{reply}

\begin{point}
    Hope authors supplement the effect of sampling and GNNs on accuracy while comparing different complexity of model and sampling methods.
\end{point}

\begin{reply}

    We are grateful for your insightful suggestion.
    %
    Followed your suggestion, we additionally evaluate the effects of model complexity on accuracy in Section 4.1 ``Effects of Hyper-parameters on Performance'' in the revised manuscript.
    %
    We add a discussion on how the hyper-parameters of GNNs affect the accuracy.
    %
    We also compare the test accuracy of different GNN models on the same dataset.

### 2020-12-26 NEUROC/sec1-introduction.tex
更改了一遍语法

To train GNNs easily, a series of GNN libraries/systems \cite{PyG, DGL, ma2019_neugraph, zhu2019_aligraph, PGL} are proposed.
%
PyTorch Geometric (PyG) \cite{PyG}, NeuGraph \cite{ma2019_neugraph}, PGL \cite{PGL} and Deep Graph Library (DGL) \cite{DGL} build upon the existing deep learning frameworks (PyG on PyTorch, NeuGraph on TensorFlow, PGL on PaddlePaddle, DGL on multiple backends).
%
They provide users with a high-level programming models (the message-passing framework for PyG/PGL/DGL and the SAGA-NN model for NeuGraph) to describe the structure of a GNN.
%
They take advantage of the common tools provided by the underlying frameworks like the automatic differentiation to simplify the development.
%
They utilize specially optimized CUDA kernels (like kernel fusion \cite{DGL, ma2019_neugraph}) and other implementation techniques (like 2D graph partitioning \cite{ma2019_neugraph}) to improve the speed of GNN training on GPUs.

@@

\begin{itemize}
    \item \emph{The training and inference time and the memory usage of a GNN layer is mainly affected by the dimensions of the input/output hidden vectors.}
    %
    Fixing other hyper-parameters, the training and inference time and the memory usage of a GNN layer increase linearly with the dimensions.

@@

    \item \emph{The high memory usage of the edge calculation stage is the main factor limiting the data scalability of GNN training and inference.}
    %
    The edge calculation generates (and caches) many intermediate results.
    %
    They are an order of magnitude larger than the dataset itself.
    %
    As GPUs have limited on-chip memory, the high memory consumption prevents us from performing training and inference on big graphs.

>>>>

To train GNNs easily, a series of GNN libraries/systems \cite{PyG, DGL, ma2019_neugraph, zhu2019_aligraph, PGL} are proposed.
%
PyTorch Geometric (PyG) \cite{PyG}, NeuGraph \cite{ma2019_neugraph}, PGL \cite{PGL} and Deep Graph Library (DGL) \cite{DGL} build upon the existing deep learning frameworks (PyG on PyTorch, NeuGraph on TensorFlow, PGL on PaddlePaddle, DGL on multiple backends).
%
They provide users with a high-level programming model (the message-passing framework for PyG/PGL/DGL and the SAGA-NN model for NeuGraph) to describe the structure of a GNN.
%
They take advantage of the common tools provided by the underlying frameworks like the automatic differentiation to simplify the development.
%
They utilize specially optimized CUDA kernels (like kernel fusion \cite{DGL, ma2019_neugraph}) and other implementation techniques (like 2D graph partitioning \cite{ma2019_neugraph}) to improve the speed of GNN training on GPUs.

@@

\begin{itemize}
    \item \emph{The training and inference time and the memory usage of a GNN layer are mainly affected by the dimensions of the input/output hidden vectors.}
    %
    Fixing other hyper-parameters, the training and inference time and the memory usage of a GNN layer increase linearly with the dimensions.

@@

    \item \emph{The high memory usage of the edge calculation stage is the main factor limiting the data scalability of GNN training and inference.}
    %
    The edge calculation generates (and caches) many intermediate results.
    %
    They are an order of magnitude larger than the dataset itself.
    %
    As GPUs have limited on-chip memory, high memory consumption prevents us from performing training and inference on big graphs.

### 2020-12-26 NEUROC/sec2-review-of-graph-neural-network.tex
更改了一遍语法

To train GNNs in a mini-batch way, the sampling techniques \cite{hamilton2017_graphsage, ying2018_pinsage, chen2018_fastgcn, chen2018_sgcn, zeng2018_aesg, chiang2019_cluster_gcn, zeng2020_graphsaint} are proposed.
%
In each mini-batch, they sample a small subgraph from the whole graph $\mathcal{G}$ and uses the subgraph to update the model parameters.
%
The sampling techniques only active the graph neurons and the connections that appear in the sampled subgraph between GNN layers, as shown in \figurename~\ref{fig:gnn_sampling}.
%
Inactive graph neurons and connections do not participate in the training of this mini-batch, saving lots of computation and storage costs.
%
Moreover, it may reduce the risk of overfitting the training graph.
%
The existing sampling techniques can be classified into two groups: \emph{neighbor sampling} and \emph{graph sampling}, based on whether different GNN layers sample different subgraphs.

>>>>

To train GNNs in a mini-batch way, the sampling techniques \cite{hamilton2017_graphsage, ying2018_pinsage, chen2018_fastgcn, chen2018_sgcn, zeng2018_aesg, chiang2019_cluster_gcn, zeng2020_graphsaint} are proposed.
%
In each mini-batch, they sample a small subgraph from the whole graph $\mathcal{G}$ and uses the subgraph to update the model parameters.
%
The sampling techniques only activate the graph neurons and the connections that appear in the sampled subgraph between GNN layers, as shown in \figurename~\ref{fig:gnn_sampling}.
%
Inactive graph neurons and connections do not participate in the training of this mini-batch, saving lots of computation and storage costs.
%
Moreover, it may reduce the risk of overfitting the training graph.
%
The existing sampling techniques can be classified into two groups: \emph{neighbor sampling} and \emph{graph sampling}, based on whether different GNN layers sample different subgraphs.

### 2020-12-26 NEUROC/sec5-insights.tex
更改了一遍语法

    \item \emph{The computational cost of a GNN layer is mainly affected by the dimensions of the input and output hidden vectors.}
          Theoretically and empirically, the training and inference time and the memory usage of a GNN layer both increase \emph{linearly} with the dimensions of the input/output hidden vectors separately.
          GNNs are friendly to high-dimensional scenarios. 
          Algorithm engineers can use high-dimensional feature vectors to improve the expressive power of a GNN without worrying exponential growth in the training/inference time and memory usage
    \item \emph{Performance optimizations should focus on improving the efficiency of the edge calculation stage.}
          The edge calculation stage is the most time-consuming stage in most GNNs.
          \begin{itemize}
              \item If the complexity of the messaging function $\phi$ is high, the implementation of $\phi$ is critical to performance.
                    Improving its efficiency can significantly reduce training/inference time.
                    For example, the attention mechanism in GNNs (like GAT and GaAN) requires an extra sub-layer to calculate the attention weight of each edge.
                    Implementing the attention mechanism with specially optimized basic operators on the GPU side is a potential optimization direction.
              \item If the complexity of $\phi$ is low, the efficiency of the collection step and the aggregation step becomes critical.
                    The existing GNN libraries \cite{DGL, PyG, ma2019_neugraph} already introduce the \emph{fused} operator to improve their efficiency.
                    When the messaging function $\phi$ is an assignment or a scalar multiplication of the input hidden vector of the source vertex, the libraries replace the collection, messaging, and aggregation steps with a single fused operator.
                    The fused operator calculates the aggregated vectors directly from the input hidden vectors, minimizing the memory footprints and overlapping the memory accessing with computation.
                    In this way, it significantly reduces the training/inference time of GNNs with low edge calculation complexity (like GCN) \cite{yan2020_characterizing_gcn, zhang2020_analysis_neugraph}.
                    However, the applicable condition of the fused operator is very restricted.
                    It does not work for $\phi$ with more complex operations like matrix multiplication.
                    A potential optimization is to develop composite CUDA kernels that can read the input hidden vectors and aggregate message vectors on the fly, without materializing the parameter vectors and the message vectors.
         \end{itemize}
    \item \emph{The high memory usage caused by the intermediate results of the edge calculation stage limits the data scalability of GNN training/inference.}
          The memory expansion ratios of the typical GNNs are very high, making GPUs unable to handle big graphs.
          One solution is to distribute the dataset among several GPUs and frequently swap parts of the dataset between GPUs and the main memory \cite{ma2019_neugraph}.
          Another possible solution \cite{chen2016_training_deep} comes from the deep neural network training. % checkpoints only work for GNN training.
          It only checkpoints key intermediate results during the forward propagation and re-calculates the missing results on demand during the backpropagation.
          Implementing the checkpoint mechanism is another potential optimization for GNN training.

>>>>

    \item \emph{The computational cost of a GNN layer is mainly affected by the dimensions of the input and output hidden vectors.}
          Theoretically and empirically, the training and inference time and the memory usage of a GNN layer both increase \emph{linearly} with the dimensions of the input/output hidden vectors separately.
          GNNs are friendly to high-dimensional scenarios. 
          Algorithm engineers can use high-dimensional feature vectors to improve the expressive power of a GNN without worrying about exponential growth in the training/inference time and memory usage
    \item \emph{Performance optimizations should focus on improving the efficiency of the edge calculation stage.}
          The edge calculation stage is the most time-consuming stage in most GNNs.
          \begin{itemize}
              \item If the complexity of the messaging function $\phi$ is high, the implementation of $\phi$ is critical to performance.
                    Improving its efficiency can significantly reduce training/inference time.
                    For example, the attention mechanism in GNNs (like GAT and GaAN) requires an extra sub-layer to calculate the attention weight of each edge.
                    Implementing the attention mechanism with specially optimized basic operators on the GPU side is a potential optimization direction.
              \item If the complexity of $\phi$ is low, the efficiency of the collection step and the aggregation step becomes critical.
                    The existing GNN libraries \cite{DGL, PyG, ma2019_neugraph} already introduce the \emph{fused} operator to improve their efficiency.
                    When the messaging function $\phi$ is an assignment or a scalar multiplication of the input hidden vector of the source vertex, the libraries replace the collection, messaging, and aggregation steps with a single fused operator.
                    The fused operator calculates the aggregated vectors directly from the input hidden vectors, minimizing the memory footprints and overlapping the memory accessing with computation.
                    In this way, it significantly reduces the training/inference time of GNNs with low edge calculation complexity (like GCN) \cite{yan2020_characterizing_gcn, zhang2020_analysis_neugraph}.
                    However, the applicable condition of the fused operator is very restricted.
                    It does not work for $\phi$ with more complex operations like matrix multiplication.
                    A potential optimization is to develop composite CUDA kernels that can read the input hidden vectors and aggregate message vectors on the fly, without materializing the parameter vectors and the message vectors.
         \end{itemize}
    \item \emph{The high memory usage caused by the intermediate results of the edge calculation stage limits the data scalability of GNN training/inference.}
          The memory expansion ratios of the typical GNNs are very high, making GPUs unable to handle big graphs.
          One solution is to distribute the dataset among several GPUs and frequently swap parts of the dataset between GPUs and the main memory \cite{ma2019_neugraph}.
          Another possible solution \cite{chen2016_training_deep} comes from the deep neural network training. % checkpoints only work for GNN training.
          It only checkpoints key intermediate results during the forward propagation and re-calculates the missing results on demand during the backpropagation.
          Implementing the checkpoint mechanism is another potential optimization for GNN training.
